---
title: "Con_bio_introduced"
output: html_document
date: "2025-01-30"
---
1. Import file
```{r}
library (tidyverse)
library (readxl)
library (dplyr)
```
2. Rename plants
```{r}
# Create a named vector for plant_name -> plant_code mapping
plant_code_map <- setNames(updated_p_and_b$plant_code, updated_p_and_b$plant_name)

# Replace nectar_species in original_matrix with corresponding plant_code
original_matrix$nectar_species <- sapply(original_matrix$nectar_species,
  function(x) plant_code_map[x])

# Result
print(original_matrix)

```

3. Rename butterflies
```{r}
# Create a named vector for plant_name -> plant_code mapping
butterfly_code_map <- setNames(updated_p_and_b$butterfly_code, updated_p_and_b$butterfly_name)

# Replace nectar_species in original_matrix with corresponding plant_code
original_matrix$butterfly_species <- sapply(original_matrix$butterfly_species,
  function(x) butterfly_code_map[x])

view(original_matrix)
```

4. Export new csv
```{r}
write.csv(original_matrix,"C:/Users/GCano/Documents/GitHub/Con_Bio_introduced/updated_matrix.csv", row.names = FALSE)
```
4. Remove NAs/Blanks
```{r}
prep_matrix<-updated_matrix
prep_matrix<- prep_matrix [!is.na(updated_matrix$nectar_species) & !is.na(updated_matrix$butterfly_species),]
```

5. Label plants as introduced/native
```{r}
introduced_plants<-c ("ELUM", "LOCO6", "SECUR3", "LONIC", "CENTA", "BRASS", "LEVU", "ROMU", "CIAR4", "LIVU2", "POTEN", "DIAR", "FRAGA", "LIGUS2", "ECVU", "DACA6", "AGERA2")

intro_matrix<-prep_matrix
intro_matrix$native_status<- ifelse(prep_matrix$nectar_species %in% introduced_plants, "introduced", "native")
```
6. Label butterflies as generalist/specialist
```{r}
general_butterflies <- c("CEPE", "CESP", "EPCL", "EUCL","EULI", "LIARC", "LIART", "LICAR", "LYPH", "MECY", "NYAN", "PAGL", "PAPI", "PAPO", "PATR", "PHSE", "PHTH", "PIRA", "POSP", "STME", "VASP")

general_matrix<-intro_matrix
general_matrix$diet_pref<-ifelse(intro_matrix$butterfly_species %in% general_butterflies, "generalist", "specialist")
```
7. Visualize
```{r}
behavior_3_data <- general_matrix %>%
  filter(Behavior == 3)

diet_proportion_data <- behavior_3_data %>%
  group_by(native_status, diet_pref) %>%
  summarise(CountOfBehavior3 = n(), .groups = "drop")  # Count how many times Behavior 3 occurred by diet preference

# Create a stacked bar plot
ggplot(diet_proportion_data, aes(x = diet_pref, y = CountOfBehavior3, fill = native_status)) +
  geom_bar(stat = "identity") +  # Use bars to represent counts
  labs(x = "Native Status", y = "Count of Behavior 3", title = "Count of Behavior 3 by Native Status and Diet Preference") +
  scale_fill_manual(values = c("native" = "lightgreen", "introduced" = "lightblue")) +  # Customize colors for diet preference
  theme_minimal()


```

```{r}
proportion_data <- general_matrix %>%
  group_by(diet_pref, native_status) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(diet_pref) %>%
  mutate(Proportion = Count / sum(Count))  # Calculate the proportion of native vs introduced within each species type

# Create a bar plot showing the proportions
ggplot(proportion_data, aes(x = diet_pref, y = Proportion, fill = native_status)) +
  geom_bar(stat = "identity", position = "fill") +  # 'fill' stacks the bars to show proportions
  labs(x = "Species Type", y = "Proportion of Native vs Introduced", title = "Proportion of Native vs Introduced Species by Generalist/Specialist") +
  scale_fill_manual(values = c("native" = "lightblue", "introduced" = "lightcoral")) +  # Customize fill colors
  theme_minimal()
```

```{r}
plant_proportion_data <- general_matrix %>%
  group_by(diet_pref, native_status) %>%
  summarise(Count = n(), .groups = "drop") %>%
  group_by(native_status) %>%
  mutate(Proportion = Count / sum(Count))

ggplot(plant_proportion_data, aes(x = native_status, y = Proportion, fill = diet_pref)) +
  geom_bar(stat = "identity", position = "dodge") +  # 'fill' stacks the bars to show proportions
  labs(x = "Species Type", y = "Proportion of Native vs Introduced", title = "Proportion of Native vs Introduced Species by Generalist/Specialist") +
  scale_fill_manual(values = c("generalist" = "lightblue", "specialist" = "lightcoral")) +  # Customize fill colors
  theme_minimal()
```

find the 5 most common introduced species using columns nectar_species and native_status from df general_matrix and then make a bar graph showing the proportion of those plants used by a generalist or specialist from the column diet_pref
```{r}
introduced_species <- general_matrix %>%
  filter(native_status == "introduced")

# Step 2: Find the 5 most common introduced species by nectar_species
top_5_introduced_species <- introduced_species %>%
  group_by(nectar_species) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(desc(count)) %>%
  head(5)  # Select the top 5 most common introduced species

# Step 3: Filter the original dataset to include only the top 5 introduced species
top_5_data <- introduced_species %>%
  filter(nectar_species %in% top_5_introduced_species$nectar_species)

# Step 4: Group by nectar_species and diet_pref, and calculate the count of occurrences for each combination
test_diet_prop <- top_5_data %>%
  group_by(nectar_species, diet_pref) %>%
  summarise(count = n(), .groups = "drop")

test_diet_prop <- test_diet_prop %>%
  group_by(nectar_species) %>%
  mutate(Proportion = count / sum(count)) 

# Step 5: Create a bar plot showing the proportion of generalists vs specialists using each introduced plant
ggplot(test_diet_prop, aes(x = nectar_species, y = Proportion, fill = diet_pref)) +
  geom_bar(stat = "identity", position = "dodge") +  # Side-by-side bars
  labs(x = "Introduced Species", y = "Count of Generalists/Specialists", 
       title = "Diet Preference for the 5 Most Common Introduced Species") +
  scale_fill_manual(values = c("generalist" = "lightblue", "specialist" = "lightcoral")) +  # Customize colors
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```

same thing for the 5 most common native species
```{r}
native_species <- general_matrix %>%
  filter(native_status == "native")

# Step 2: Find the 5 most common native species by nectar_species
top_5_native_species <- native_species %>%
  group_by(nectar_species) %>%
  summarise(count = n(), .groups = "drop") %>%
  arrange(desc(count)) %>%
  head(5)  # Select the top 5 most common native species

# Step 3: Filter the original dataset to include only the top 5 native species
top_5_native_data <- native_species %>%
  filter(nectar_species %in% top_5_native_species$nectar_species)

# Step 4: Group by nectar_species and diet_pref, and calculate the count of occurrences for each combination
diet_proportion_native_data <- top_5_native_data %>%
  group_by(nectar_species, diet_pref) %>%
  summarise(count = n(), .groups = "drop")

diet_proportion_native_data <- diet_proportion_native_data %>%
  group_by(nectar_species) %>%
  mutate(Proportion = count / sum(count))

# Step 5: Create a bar plot showing the proportion of generalists vs specialists using each native plant
ggplot(diet_proportion_native_data, aes(x = nectar_species, y = Proportion, fill = diet_pref)) +
  geom_bar(stat = "identity", position = "dodge") +  # Side-by-side bars
  labs(x = "Native Species", y = "Count of Generalists/Specialists", 
       title = "Diet Preference for the 5 Most Common Native Species") +
  scale_fill_manual(values = c("generalist" = "lightblue", "specialist" = "lightcoral")) +  # Customize colors
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

